Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-03-28 VAST_submission/vgtc_journal_latex/paper.tex
Added some references, updated and added to the paper

In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, displaying simple characteristics of the emails,  the so-called metadata.

@@

\TODO{This needs to be short and to the point.  Just say what is done.}
Before any of the analysis could begin, the emails had to be converted from the individual PDF and HTML form into a more easily used data structure. The choice was made to use the Wikileaks HTML emails as the source for data extraction. This choice was made primarily due to the relative ease of loading the raw HTML pages on the Wikileaks database due to their regular method of storage (the email with ID <id> is stored at https://wikileaks.org/clinton-emails/emailid/<id>) and ease of loading. These qualities made the HTML emails amenable to programmatic extraction, which was completed using a host of web-scraping and string parsing tools in \texttt{R}. In particular the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form. After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.

Unfortunately, this data is not perfect. The PDF files released by the state department do not include the typical email server header with information about the time stamp inforamtion, the addresses of all involved, and other useful and regular information. Instead, these files appear more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display. As a result, the email addresses are occasionally not included, and there were initially concerns over the time zones used to generate the time stamp metadata. Furthermore, the HTML data on Wikileaks is frequently irregular in form, and although work has been done to make the cleaning and extraction functions as general as possible, there are still certainly fringe cases not considered which occur frequently enough to have an impact. It is also important to keep in mind the fact that the data have already been filtered twice. First by the emails chosen to be released by Clinton, and next by the emails and email sections which the State Department chose not to redact. Indeed, even the metadata can be affected by this redaction, as a number of individuals have their email addresses redacted in all communications, preventing certain analysis from being performed.

\section{Displays}
\TODO{Each of these four displays is described and commented on in context of whole time line}
\subsection{Inner circle} 
The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals. As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction. Finally, all edges and points are coloured according to a simple scheme. If the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, they are coloured blue. Individuals with email addresses which are identifiably not hosted at a '.gov' domain are coloured red. Finally, individuals which cannot have their email host domain identified are coloured grey. It is important to note that for those individuals with grey spokes, the reason the email is not identifiable is the removal of the email by the state department. Such a redaction is not a guarantee that the email said individual used is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain. Still, such speculation cannot be conclusively made, and so these emails are simply reported as they are, unidentifiable.

@@

\subsection{FOIA exemptions}.
 It is important to notice that this barplot is not trivialized by the possibility of selecting individual FOIA exemption codes, as many of the emails in the data set are exempted on a number of grounds, and so have numerous FOIA exemption codes. Thus, this barplot is not only interesting when all codes are selected, but can be used to see which codes commonly co-occur in documents.

\subsection{Term frequency and TFIDF}

\section{Filters}
\TODO{Each filter is described and displayed }
\subsection{Time filtering}
\subsection{Sent or received}
\subsection{FOIA exemptions}
\subsection{Auxiliary information}
\TODO{e.g. Clinton's foreign schedule}
\section{Some further exploration}
\TODO{Some of our findings.  Unless done in the previous section}
\section{Shiny App}
\TODO{Again short and to the point}
The app (HOST ADDRESS) provides the ability for users to place custom filters on the time frame of server time stamps and classification tags of the emails and view how a series of three displays change.

Finally, a barplot of the FOIA exemption codes used in the selected data is displayed

>>>>

In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, the so-called metadata. The reasons for this are two-fold. The first is that for this data the metadata is simply more reliable and regular. The emails are often heavily redacted, and this introduces both gaps in content and irregularities. The metadata, however, remains somewhat informative and regular in format despite redactions, as the only redacted metadata is that of individual email addresses, and noting these redactions is in itself useful. The second reason is demonstrate the power of supported metadata in the exploration of communication data. For public figures in particular, a communication network plot and time-stamps can be very easily leveraged for insightful data exploration if outside sources are consulted. The individual's communication habits can be extended to gain insight into their general habits in life, while the individuals they communicate with most frequently can establish interesting connections which might otherwise be missed, such as that between Clinton and Tony Blair.

@@

\TODO{This needs to be short and to the point.  Just say what is done.}
HTML versions of the emails were pulled programmatically from the Wikileaks database, as the storage structure supported programmatic extraction using web-scraping and string parsing tools in \texttt{R}. In particular, the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form. After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.

The PDF files released by the state department used to generate the Wikileaks HTML emails do not include the typical email server header with time stamp information, email addresses, and other regular data. Instead, these files are more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display. As a result, the data is very irregular: email addresses are occasionally not included, there were initially concerns over the time zones used to generate the time stamp metadata, and there are doubtless still many fringe cases of irregular spacing or placement of text within the HTML code not considered in cleaning. While the metadata is relatively consistent in spite of these irregularities, there are still impure and messy addresses and contact names arising from these irregularities.

After the emails were pulled, stopwords defined by the \texttt{stopwords} function from \texttt{tm} were removed and the remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}. Both of these seemed imperfect and ill-suited to the data, failing to remove words with little informative power and creating often cryptic stems.

\section{Displays}
\TODO{Each of these four displays is described and commented on in context of whole time line}
\subsection{Inner circle} 
The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals. As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction. Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange. It is important to note that for those individuals with orange spokes, the most likely reason the email is not identifiable is the redaction of the email by the state department. Such a redaction is not a guarantee that the email of the individual is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.

@@

One such result which is immediately obvious is the strange pattern of sending times. Email communications occur with the highest frequency between midnight and 3 am, and with the lowest frequency between 4 pm and 10 pm. This pattern, which is persistent through filters of emails sent to Clinton and by Clinton, is surprising as it indicates that the highest level of communication was occurring at a time typically devoted to sleep. Another pattern immediately obvious is the alternating modes at 2 and 3 am, with alternations occurring on the dates of daylight savings in North America. The regularity of this pattern in the emails suggests some server function performed early in the morning, perhaps some maintenance operation performed every 24 hours.

\subsection{FOIA exemptions}.
 The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request. The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions. The frequencies of each of these codes within filtered emails are shown in the final display. Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.

\subsection{Term frequency and TFIDF}
The term frequency and tf-idf displays for this data are perhaps the least directly useful. As mentioned above, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} proved inadequate for the messy and non-standard data. In particular, the omission of many common and uninformative action verbs such as 'will' and 'can' from the stopwords used led to very unsatisfactory results. A set of custom stopwords was therefore applied in an attempt to remove some of these uninformative words. The tf-idf was more informative, but the stemming algorithm, when coupled with the typos and abbreviations which are incredibly common in the emails, meant that the results are still incredibly unclear. The terms with the highest scores over any interval do not seem to communicate anything particularly useful about the content of emails sent in that range, with one notable exception in the case of the term 'windrush.' This term appeared with a high tf-idf in a number of selections which contained communications with Tony Blair (email contact "aclb" in the spiral plot), and seems to refer to Windrush Ventures, a company owned by Blair which appears vital in managing Blair's complicated and opaque finances since he left the office of the Prime Minister of the United Kingdom. \cite{windrushTelegraph, windrushGuardian}

\section{Filters}
\TODO{Each filter is described and displayed }
\subsection{Time filtering}
Time filtering within the app can be completed at the level of individual days. The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days. As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width. Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
\subsection{Sent or received}
A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported in the app with a box selection area. Users can add FOIA exemption codes currently not included to the data by clicking the box with their mouse and selecting a code from the menu which is then displayed below the box. To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu. The box therefore displays all codes currently included in the data displayed.
\subsection{Auxiliary information}
\TODO{e.g. Clinton's foreign schedule}
A number of cosmetic adjustments using auxiliary information can also be made. Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes. This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic. Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot. This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).
\section{Some further exploration}
\TODO{Some of our findings.  Unless done in the previous section}
\section{Shiny App}
\TODO{Again short and to the point}
The \texttt{R} \texttt{shiny} app at (HOST ADDRESS) provides the ability for users to interactively filter the data on all features described above and view the resulting displays. Alongside this display is a short article explaining the use of the app, providing links to background information, and outlining some of the more interesting trends observed. It is hosted on the Waterloo \texttt{rshiny} server for public access, allowing for any individual to make use of this tool to generate hypotheses and investigate this intriguing dataset.
